Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-04-03 paper.tex
minor writing edits in intro

In \S\ref{setup} we list the governing equations and global
equilibrium disk models under   
consideration. We formulate the linear stability problem in 
\S\ref{linear} and simplify it by making the radially local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work and 
allows us to highlight the essential effects of vertical shear, and to
derive a characteristic upper limit to the thermal timescale for the
VSI. In \S\ref{numerical} we numerically
solve the linear problem in the radially local approximation, but
relax all other approximations made in our analytical approach.    
We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale.

@@

\section{Thermodynamic requirement for the vertical shear instability}\label{vsi_require}    
In this paper we will find that in a stably stratified disk, the VSI
requires a \emph{very} short thermal timescale $t_c$, much less than the
disk dynamical timescale,
\begin{align}
  t _c \ll \Omega^{-1},  
\end{align}
consistent with numerical experiments \citep{nelson13}. For this
preliminary discussion we shall consider a vertically isothermal disk
with aspect-ratio $\epsilon$, 
dimensionless radial temperature gradient $q$ and adiabatic index
$\gamma$.

In a thin disk, the specific angular momentum varies with height as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega^2 }{\p z} \simeq \epsilon q \Omega^2\left(\frac{z}{H}\right),
\end{align}
where $H$ is the disk scale height. As noted in
\S\ref{intro} this vertical variation is a source of free energy and
thus vertical shear ($\p_z\Omega \neq0$) is destabilizing.

If $\gamma>1$ then the disk is stably stratified with a real vertical
buoyancy frequency $N_z$.  For a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \Omega^2.  
\end{align}
This provides a stabilizing effect as buoyancy opposes vertical
motion, which is required for the VSI. In fact, if we consider
axisymmetric adiabatic disturbances, then because the 
Solberg-Hoiland criteria \citep{tassoul78} are usually satisified in a
thin disk, the disk would be stable.

However, the stabilizing effect of buoyancy can be mitigated for
non-adiabatic disturbances. Consider a fluid element displaced upwards, away from
the disk midplane. In a stably stratified adiabatic disk, it would
sink because of buoyancy. However, if thermal timescales are short, in
that temperature flucations decay rapidly (i.e. nearly isothermal
perturbations), then the fluid element would attain its surrounding
density and  
%(assuming an ideal gas and it reaches pressure 
%equilibrium with its surroundings). 
thus buoyancy is rendered ineffective. Qualitatively, a short thermal
timescale $t_c$ reduces the effective (stabilizing) buoyancy force.

How short must $t_c$ be for vertical shear to prevail? For a rough
comparison of vertical shear to buoyancy for non-adiabatic
perturbations, we apply Eq. 47 of \cite{urpin03}, which is based on
\cite{townsend58},   
\begin{align}\label{bcrit_est}
  N_z^2 t_c \lesssim \left|r\frac{\p\Omega}{\p z}\right|. 
\end{align}
The left-hand side represents an effective buoyancy, which is
stabilizing; while the right-hand side represents vertical shear,
which is destabilizing. In a thin disk, the destabilizing effect of
vertical shear is rather weak because $|r\p_z\Omega|\propto \epsilon
\ll 1$. By comparison the buoyancy frequency $N_z/\Omega\sim 1$ at
$z\sim H$.

To illustrate this point, we show in Fig. \ref{eqm_structure} the
equilibrium structure of a typical protoplanetary disk. Here $\rho_0$
is the midplane density and vertical shear rate and buoyancy
frequencies are normalized by the local Keplerian frequency. It is
clear that vertical shear is 
generally weak compared to buoyancy. Then even a small (positive)
vertical entropy gradient can significantly stabilize the VSI. It
follows that $t_c$ must be quite short to permit instability.

@@

Our central result is the requirement 
\begin{align}\label{prelim_bcrit}
  (\gamma - 1)t_c < \epsilon |q| \Omega^{-1}
\end{align}   
for the VSI in a vertically isothermal, thin disk. This equation
reflects Eq. \ref{bcrit_est} but is simpler since there is no
$z$ dependence. \emph{In a thin 
  disk the thermal timescale must be shorter than the dynamical
  timescale by $O(\epsilon)\ll 1$ for the VSI to operate.}

% While assumptions and simplifications will be made 
% to derive this,

Eq. \ref{prelim_bcrit} can be 
readily applied to evaluate where in a disk the VSI is likely to
operate. For example, we show in Fig. \ref{bcrit_mmsn2d} an estimate
of the thermal timescales the protoplanetary disk model shown in
Fig. \ref{eqm_structure}. Here $k_x$ is a radial
wavenumber of the disturbance being considered, thus larger $k_x$
correspond to smaller perturbations. The region outside the red line
satisfies Eq. \ref{prelim_bcrit}. The criterion Eq. \ref{prelim_bcrit} in fact
assumes $t_c$ is vertically uniform. While this is not generally true
in the inner disk, we find Eq. \ref{prelim_bcrit} still provides a
useful guide. We see that the VSI will operate in the outer 
disk at tens of AU, but that smaller scale disturbances may operate at
smaller radii.

>>>>

In \S\ref{setup} we list the governing equations and global
equilibrium disk models under   
consideration. We formulate the linear stability problem in 
\S\ref{linear} and simplify it by making the radially local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work, 
allowing us to highlight the essential effects of vertical shear, and to
derive a characteristic upper limit to the thermal timescale for the
VSI. This timescale is \emph{much shorter} than the dynamical
timescale. In \S\ref{numerical} we numerically
solve the linear problem in the radially local approximation, but
relax all other approximations made in our analytical approach.    
We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale.

@@

\section{Thermodynamic requirement for the vertical shear instability}\label{vsi_require}     In this paper we will find that in a stably stratified thin disk, the VSI  
requires a thermal timescale $t_c$ much less than the
disk dynamical timescale, 
\begin{align}
  t _c \ll \Omega^{-1}. 
\end{align}
To see why this should be expected, we need to compare vertical shear
with vertical buoyancy. For this preliminary discussion we consider a
vertically isothermal disk with aspect-ratio $\epsilon$,  
dimensionless radial temperature gradient $q$ and adiabatic index
$\gamma$. For a protoplanetary disk, $\epsilon \sim 0.05$, $|q|\sim 1$ and $\gamma\simeq 1.4$.

For a thin disk, the specific angular momentum varies with height as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega^2 }{\p z} \simeq \epsilon q \Omega^2\left(\frac{z}{H}\right),
\end{align}
where $H$ is the disk scale height. As noted in
\S\ref{intro} this vertical variation is a source of free energy and
thus vertical shear ($\p_z\Omega \neq0$) is destabilizing.

For $\gamma>1$ the disk is stably stratified with a real vertical
buoyancy frequency $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \Omega^2.  
\end{align}
This provides a stabilizing effect as buoyancy opposes vertical
motion, which is required for the VSI. In fact, if we consider
axisymmetric adiabatic disturbances, then because the 
Solberg-Hoiland criteria \citep{tassoul78} are usually satisified in a
thin disk, the disk would be stable.

However, the stabilizing effect of buoyancy can be mitigated for 
effectively isothermal disturbances. Consider a fluid element displaced upwards, away from
the disk midplane. In a stably stratified adiabatic disk, it would
sink because of buoyancy. However, if thermal timescales are short, in
that temperature flucations decay rapidly, then the fluid element would attain its surrounding 
density and  
%(assuming an ideal gas and it reaches pressure 
%equilibrium with its surroundings). 
thus buoyancy is rendered ineffective. Qualitatively, a short thermal
timescale $t_c$ reduces the effective (stabilizing) buoyancy force.

How short must $t_c$ be for vertical shear to prevail? A rough 
comparison of vertical shear to buoyancy for non-adiabatic
perturbations can be obtained by applying Eq. 47 of \cite{urpin03}, which is based on 
\cite{townsend58},   
\begin{align}\label{bcrit_est}
  N_z^2 t_c \lesssim \left|r\frac{\p\Omega}{\p z}\right|. 
\end{align}
The left-hand side represents an effective buoyancy, which is
stabilizing; while the right-hand side represents vertical shear,
which is destabilizing. In a thin disk $\epsilon \ll 1$, so the destabilizing effect of 
vertical shear is rather weak because $|r\p_z\Omega|\sim\epsilon\Omega
\ll \Omega$. By comparison, the buoyancy frequency $N_z\sim
\Omega$. It follows that $t_c\lesssim O(\epsilon\Omega^{-1})$ is
required for instability.  \emph{In a thin  
  disk the thermal timescale must be shorter than the dynamical
  timescale by $O(\epsilon)\ll 1$ for the VSI to operate.}

To illustrate this point, we show in Fig. \ref{eqm_structure} the
equilibrium structure of a typical protoplanetary disk. Here $\rho_0$
is the midplane density and vertical shear rate and buoyancy
frequencies are normalized by the local Keplerian frequency. It is
clear that vertical shear is 
generally weak compared to buoyancy. Then even a small (positive)
vertical entropy gradient can significantly stabilize the VSI.

@@

Our central result is the requirement 
\begin{align}\label{prelim_bcrit}
  (\gamma - 1)t_c < \epsilon |q| \Omega^{-1}
\end{align}   
for the VSI in a vertically isothermal, thin disk. This equation
reflects Eq. \ref{bcrit_est} but there is no $z$ dependence. 
% While assumptions and simplifications will be made 
% to derive this, 
Eq. \ref{prelim_bcrit} can be 
readily applied to evaluate where in a disk the VSI is likely to
operate. For example, we show in Fig. \ref{bcrit_mmsn2d} an estimate
of the thermal timescales the protoplanetary disk model shown in
Fig. \ref{eqm_structure}. Here $k_x$ is a radial
wavenumber of the disturbance being considered, thus larger $k_x$
correspond to smaller perturbations. The region outside the red line
satisfies Eq. \ref{prelim_bcrit}.

The criterion Eq. \ref{prelim_bcrit} in fact
assumes $t_c$ is vertically uniform. While this is not generally true
in the inner disk, we find Eq. \ref{prelim_bcrit} still provides a
useful guide. We see that the VSI will operate in the outer 
disk at tens of AU, but that smaller scale disturbances may operate at
smaller radii.
